---
title: "Reliability Analysis"
author: "Ana Milovanović / Martí Bernardo /Edited by: Raquel Álvarez-Urdiola"
date: "2024-09-30"
output: pdf_document
toc: TRUE
editor_options: 
  chunk_output_type: console
---

# Reliability Analysis
## Identify reliable days
Neighbours are the immediately adjacent days to the day of interest, i.e. the day before or the day after. One reliable neighbour is enough to "save the day", i.e. for the day not to be classified as an NA.

Reliable neighbours discussion  

* Day 0 - has only one neighbour, Day 1, which is its reliable neighbour if it has <=2 NAs
* Day 1 - Day 0 is its reliable neighbour if it has <=2 NAs (a less stringent option, could try <=1), Day 2 is its reliable neighbour if it has <=2 NAs
* Day 5 - has only one neighbour: Day 4, which is its reliable neighbour if it has <=2 NAs
* all other days have two neighbours, each one of them a reliable neighbour if it has <=2 NAs.


Day reliability and action taken on the data:

* reliably undetected (RU) - if it has all replicates as NAs (3 NAs in Day 0 or 4 NAs in Days 1-5) and has at least 1 unreliable neighbour (but at the moment we are not checking the neighbour for D0 and D5); Action: in this day, all of the samples will be replaced with minimum of detection
* reliably detected (RD) - if it is reliable on its own (number of NAs in such a day is <=1), in which case its neighbours are not important; Action: data left as is
* unreliably detected (UD) - more-less reliable, if it is unreliable on its own but has at least one reliable neighbour; Action: data left as is
* unreliably undetected (UU) - if (i) neither reliable on its own nor has even one reliable neighbour or (ii) has all replicates as NAs and both neighbours are reliable (included (ii) to avoid dips); Action: in this day, in both cases (i) and (ii), all of the samples will be replaced with NA

Warning:
It is still possible to end up with a dip, when a day in question has all replicates as NAs and the neighbouts are unreliable, because those nbeighbours might still be deemed Unreliable Detections (UD), because of the reliability of their other neighbours.


Reliability matrix (proteins with their corresponding number of NA replicates in a day) is shown as a heatmap in Fig. \ref{fig:reliability_matrix}.
 

```{r,eval = TRUE, include = TRUE, warning=FALSE, message=F, echo = F, fig.asp=0.8, fig.cap = "Proteins with their corresponding number of NA replicates in each day (Reliability matrix).\\label{fig:reliability_matrix}"}
num_prots = dim(total.log.norm.23.no.NA.R1D0)[1]
num_days=6
##print(num_prots)
##print(num_days)
warning("The total prot num is now redefined after removing those only present in R1D0")
Reliability_matrix = matrix(nrow = num_prots, ncol = num_days)

#****************************************************
# Identify reliable Days 0 - 5
#****************************************************

#Day is reliable if NA <=1


# ALL days
for(prot_index in 1:num_prots){
  #print(paste("protein",prot_index,"---------------------"))
  
  # vector of logical values, saying whether or not the NA_by_day low enough to be  considered a good neighbour:  NA_by_day>=2
  good_neighbour = c(NA_by_day[prot_index,1]<=2, NA_by_day[prot_index,2]<=2, NA_by_day[prot_index,3]<=2, NA_by_day[prot_index,4]<=2, NA_by_day[prot_index,5]<=2, NA_by_day[prot_index,6]<=2)
  # it's a function of day_index! so, it will not be used if all days are reliable on their own, i.e. RD

  # loop over all 6 days
  for(day_index in 1:num_days){ # (num_days-1) MUST be in brackets! otherwise R also subtracts 1 from the initial value of i!!!!
    
    # broadly divide into 3 groups, the if statement, according to how many NAs there are in a day:
    
    # - Reliably Detected ----------------------------------------   
    # if up to 1 replicate is an NA  
    if (NA_by_day[prot_index, day_index] <= 1){Reliability_matrix[prot_index,day_index] = 3} # {RD}
    #........................................................
    # if >=2 (2-4) replicates are NAs
    else {
      # if 4 replicates are NAs 
      if (NA_by_day[prot_index, day_index] == 4) {
        # if days are D1-D4, not D5   (D0 cannot be here by definition, as it only has 3 replicates!!!!)
        if (day_index!=6) {  # 
          # if both neighbours are reliable
          if (good_neighbour[day_index-1] & good_neighbour[day_index+1]) {Reliability_matrix[prot_index,day_index] = 1}  # {UU} - NA
          # if it has at least 1 UNRELIABLE neighbour (one or both)! 
          else {Reliability_matrix[prot_index,day_index] = 0}  # {RU} - min_detection  
        }
        # can only be D5!!!! if all 4 replicates are NAs in D5, because it only has one neighbour anyway????
        else {Reliability_matrix[prot_index,day_index] = 0}  # {RU} - min_detection 
      }
      #........................................................
      # if 2-3 replicates are NAs (in any day? i.e. 1-2 replicates are present, except that in the case of D0 this means 0-1 replicates are present!) : RU possible for D0 if 3 NAs; for all days possible UU or UD: 
      else {   
        # D0 when it has all 3 NAs : RU case
        #print(day_index)
        #print(NA_by_day[prot_index,day_index-1])
        if (day_index==1 & NA_by_day[prot_index,day_index]==3) {Reliability_matrix[prot_index,day_index] = 0}  # {RU} - min_detection      
        # for all days, when they have 2-3 NAs: options now are UD or UU
        else {
          # day has at least one good neighbour?
          if (day_index==1) {crit_good_neighbours=good_neighbour[day_index+1]}  # D0
          if (day_index==6) {crit_good_neighbours=good_neighbour[day_index-1]}  # D5
          if (day_index>=2 & day_index<=5) {crit_good_neighbours = (good_neighbour[day_index-1] | good_neighbour[day_index+1])} # D1-4
          
          # Unreliably Detected --------------------------------------------------
          if (crit_good_neighbours) {Reliability_matrix[prot_index,day_index] = 2}    # {UD}    }
          # Unreliably Undetected --------------------------------------------------
          # Day not reliable and neighbours are not reliable
          else {Reliability_matrix[prot_index,day_index] = 1}       #  {UU} - NA   }
        }
      }
    }
    
    
     # Reliably Undetected -------------------------------------------------------
      # CHANGE introduced! to avoid dips:
    
      # when a day has all replicates as NAs (3 in D1 or 4 in all the rest) it is:
      # - Reliably Undetected (min_detection): ONLY if ONE of the 2 neighbours as Unreliable; 
      # - Unreliably Undetected (NA): if BOTH neighbours are Reliable
      # old:
      # Day has all as NAs
      # define the criterion, depending on the day:
    
      # ?????  D1 and D5 have only 1 neighbour - leave them out? apply this only to D1-4
  }
}



#print (Reliability_matrix)
# save("Reliability_matrix", file="Reliability_matrix.RData")
##View(Reliability_matrix)
##summary(Reliability_matrix)
##head(Reliability_matrix)
##colSums(which(Reliability_matrix==0))   - gives Error in colSums(which(Reliability_matrix == 0)) :   'x' must be an array of at least two dimensions

# LATER UNCOMMENT!  
#library(NMF) 
#aheatmap(Reliability_matrix, Rowv=F, Colv=NA)
source("plot_heatmap_dis.R")

colnames(Reliability_matrix) = c("Day_0", "Day_1", "Day_2", "Day_3", "Day_4", "Day_5")  # same as in NA_by_day
  
plot_heatmap_dis(Reliability_matrix, "Reliability_matrix.eps", num_categories=4)
#dev.off()

#hist(Reliability_matrix)
#****************************************************

# Select reliable Days 1-4 independent of neighbours
#****************************************************
#If NA_num_day <=1 {day_reliable ==T }

    
#Questions
#Neighgours only adjacent or also more than one day away
#Day 0 is unreliable neighbour with same threshold as other days, and we should be careful about that because it has one less replicate from the start
#Day 0 and 5 can only have one neighbour, hence with current criteria of two AND adjancent required neighbours, they would considered "neighbourless"
#Are we happy with removing 1885 proteins out of 8924 (after removing 165 that only were present in R1D0)

#ToDo 
#If NA = 100% --> unreliable, regardless  of neighbours
#-->Hence write in reliability matrix "All Replicates NA", before starting (undetected)
#If day is reliable by itself 
#-->Hence write in reliability matrix "Self-Reliable"" (consistently detected)
#If Identifies as reliable thanks to neighbours
#-->Write in matrix "Reliable by neighbour" (detected)
#If unreliable in itself and by neighbours
#-->matrix: consistently undetected
```



## Confirm that a protein with all days considered unreliable by our criteria fits the manual criteria for unreliable protein



```{r,eval = TRUE, include = TRUE, warning=TRUE, message=F, echo = F, fig.cap = ""}
# NOTE: warning=TRUE because we want to print the text that shows how many unreliable proteins there are out of the total.


#****************************************************
# Double check if we correctly identify unreliable proteins i.e. based on Unreliability_matrix
#****************************************************
#Identify on reliabiltiy matrix those proteins with 0 i.e. ureliable



#****************************************************
# Identify unreliable proteins "manually" i.e based on NA_by_day
#****************************************************
Unreliable_proteins = vector(length = num_prots)
for(prot_index in 1:num_prots){     #if more than 1 NA per day
  if(NA_by_day[prot_index,1]>=2 & 
     NA_by_day[prot_index,2]>=3 &
     NA_by_day[prot_index,3]>=3 & 
     NA_by_day[prot_index,4]>=3 &
     NA_by_day[prot_index,5]>=3 & 
     NA_by_day[prot_index,6]>=3) {
    Unreliable_proteins[prot_index]=T
  }
}

warning("The number of unreliable proteins (having more than 1 NA in every day) is ", sum(Unreliable_proteins)," out of ", num_prots, " in total.") # 1885 out of 8924 
##View(NA_by_day[4475,])


# 
# prot_index=1
# Reliability_matrix[prot_index,]==rep(0,6)
# Unreliable_test_proteins = rep(0,num_prots)
# head(Unreliable_test_proteins)
# for(prot_index in 1:num_prots){Unreliable_test_proteins[prot_index]=all(Reliability_matrix[prot_index,]==rep(0,6))}
# 
# names(Unreliable_test_proteins)=rownames(total.log.23.No.NA.R1D0)
# names(Unreliable_proteins)=rownames(total.log.23.No.NA.R1D0)
# length(intersect(names(Unreliable_proteins),names(Unreliable_test_proteins)))
# 
# 
# #****************************************************
# # do the sets of prots intersect?
# #****************************************************
# warning("Unreliable proteins: ", sum(Unreliable_proteins)," out of ", num_prots)
# 
# intersect(Unreliable_proteins_matrix, Unreliable_proteins_manual)

```



## Create a new data matrix with NAs and min_detection

Values that are zeros in the dataset are replaced with the minimum of detection and NAs

```{r,eval = TRUE, include = TRUE, warning=FALSE, message=F, echo = F, fig.cap = ""}
#****************************************************
# Replace by NA, save in dataset total.log.23.No.NA.R1D0.Reliable
#****************************************************
#(i)  RU, realiably undetected replaced by 0s 
#(ii) UU, unreliably undetected replaced by NAs (according to Reliability_matrix) the values in total.log.23.No.NA.R1D0, thereby creating total.log.23.No.NA.R1D0.Reliable

Day_0 = c(1,2,3)
Day_1 = c(4:7)
Day_2 = c(8:11)
Day_3 = c(12:15)
Day_4 = c(16:19)
Day_5 = c(20:23)

Days = list(Day_0 = Day_0, Day_1 = Day_1, Day_2 = Day_2, Day_3 = Day_3, Day_4 = Day_4, Day_5 = Day_5)

total.log.norm.23.no.NA.R1D0.Reliability = total.log.norm.23.no.NA.R1D0
# write.csv(total.log.norm.23.no.NA.R1D0.Reliability, file = "total.log.norm.23.no.NA.R1D0.Reliability.csv")

num_prots = dim(total.log.norm.23.no.NA.R1D0.Reliability)[1]

undetected_protein = vector(mode = "logical",length = num_prots)
names(undetected_protein) = rownames(total.log.norm.23.no.NA.R1D0.Reliability)
undetected_protein[] = TRUE
##head(undetected_protein)

#number_of_detections = 
counter_days = vector(length = num_prots)
names(counter_days) = rownames(total.log.norm.23.no.NA.R1D0.Reliability)
##counter_days

#min_data = min(na.omit(total.23.log.norm)) # minimum detection   ???????
min_data = min(total.log.norm.23.no.NA.R1D0.Reliability[!is.na(total.log.norm.23.no.NA.R1D0.Reliability)])  # normalised minimum detection!!!
min_detection = trunc(min_data)

num_days = 6
for(prot_index in 1:num_prots){
  counter_days[prot_index] = 0
  for(day_index in 1:num_days){
    
   if(Reliability_matrix[prot_index,day_index]==1){total.log.norm.23.no.NA.R1D0.Reliability[prot_index, Days[[day_index]]] = NA}
    #if(Reliability_matrix[prot_index,day_index]=="UU"){total.log.23.No.NA.R1D0.Reliable[prot_index, Days[[day_index]]] = NA}
   else{
     if(Reliability_matrix[prot_index,day_index]==0){total.log.norm.23.no.NA.R1D0.Reliability[prot_index, Days[[day_index]]] = min_detection}  # 0}
     #if(Reliability_matrix[prot_index,day_index]=="RU"){total.log.23.No.NA.R1D0.Reliable[prot_index, Days[[day_index]]] = 0}
     else {
       undetected_protein[prot_index] = FALSE
      counter_days[prot_index]=counter_days[prot_index]+1    # counts in how many days detected
     }
   } 
  }
  #?protein_reliable_days =  # protein and its number of reliable detections (days) 
}

##summary(undetected_protein)
##View(counter_days)

```

\clearpage


## Quantification of undetected proteins  

Fig. \ref{fig:count_detected_vs_undetected} illustrates that how many proteins are detected in at least one day vs the undetected ones, i.e. the ones that have NAs or minimum of detection in all days.



```{r,eval = TRUE, include = TRUE, warning=TRUE, message=F, echo = F, fig.align="center", out.width="85%", fig.asp=0.5, fig.cap = "The figure shows the number of proteins detected in at least one day vs the ones that are undetected, i.e. replaced either by an NA or minimum of detection in all days.\\label{fig:count_detected_vs_undetected}"}
# print the warning 

#dev.new()
par(mar=c(3,2,2,2)+0.1) # sets the bottom, left, top and right margins respectively of the plot region in number of lines of text. 


undetected_protein1=undetected_protein
undetected_protein1=replace(undetected_protein1, undetected_protein1=="FALSE", "detected")
undetected_protein1=replace(undetected_protein1, undetected_protein1=="TRUE", "undetected")


barplot_detect_vs_undetect = barplot(table(undetected_protein1))

barplot_detect_vs_undetect



warning("The number of detected proteins is ",length(which(undetected_protein1=="detected"))," out of ", length(undetected_protein1), " in total.") # 7033 out of 8924 


```

\clearpage

## Number of proteins by the number of detections  

Fig. \ref{fig:num_prots_vs_num_days_detected} shows how many proteins are detected in a particular number of days.

```{r,eval = TRUE, include = TRUE, warning=FALSE, message=F, echo = F, fig.align="center", out.width="75%", fig.asp=0.5, fig.cap = "A histogram of the number of proteins detected in a particular number of days.\\label{fig:num_prots_vs_num_days_detected} "}
par(mar=c(3,2,2,2)+0.1) # reduces the bottom margin on the plot
# 
# aheatmap(NA_by_day.Reliability, Rowv=F, Colv=NA)

barplot(table(counter_days))


# TODO: make a histogram: how many proteins detected on 0,1,2,...6 days

```



## Removal of the proteins that have all days as non-detections - NAs or min_detection (dataset total.log.norm.23.no.NA.R1D0.Reliability.no.undetected)



```{r,eval = TRUE, include = TRUE, warning=FALSE, message=F, echo = F, fig.cap = ""}
# 

prots_all_absent_1 = vector(length = dim(total.log.norm.23.no.NA.R1D0.Reliability)[1],mode = "integer")
names(prots_all_absent_1) = row.names(total.log.norm.23.no.NA.R1D0.Reliability)
for(prot_index in 1:dim(total.log.norm.23.no.NA.R1D0.Reliability)[1]){
 if( counter_days[prot_index] == 0){prots_all_absent_1[prot_index] = 1} # if undetected in ALL 6 days
}

##head(prots_all_absent_1)
##View(prots_all_absent_1)
#length(prots_all_absent_1) # 8924

warning("Number of proteins with all days as NAs:", length(which(prots_all_absent_1==1))) # 1891 

# write.csv(which(prots_all_absent_1==1),  file = "undetected_in_all days.csv") # will write the protein ID and the row number in the original table



#source("assessment_of_NA_or_0_by_day.R")
#NA_by_day.Reliability=assessment_of_NA_or_0_by_day(total.23.log.norm)
#NA_by_day.Reliability=assessment_of_NA_or_0_by_day(total.23.log)
#raquel.NAs <- as.data.frame(NA_by_day.Reliability)
#total.NAs <- mutate(raquel.NAs,
#                    total_NAs = (Day_0 + Day_5 + Day_4 + Day_3 + Day_2 + Day_1))
#head(NA_by_day.Reliability)
#for(prot_index in 1:dim(total.log.norm.23.no.NA.R1D0.Reliability)[1]){
# if( NA_by_day.Reliability[prot_index,1] == 3 & NA_by_day.Reliability[prot_index,2] == 4 &  NA_by_day.Reliability[prot_index,3] == 4 & NA_by_day.Reliability[prot_index,4] == 4 & NA_by_day.Reliability[prot_index,5] == 4 & NA_by_day.Reliability[prot_index,6] == 4){prots_all_absent_1[prot_index] = 1}
#}

# no undetected means undetected in all days!
total.log.norm.23.no.NA.R1D0.Reliability.no.undetected = total.log.norm.23.no.NA.R1D0.Reliability[-which(prots_all_absent_1==1),]
# write.csv(total.log.norm.23.no.NA.R1D0.Reliability.no.undetected, file = "total.log.norm.23.no.NA.R1D0.Reliability.no.undetected.csv")
# save(total.log.norm.23.no.NA.R1D0.Reliability.no.undetected, file = "total.log.norm.23.no.NA.R1D0.Reliability.no.undetected.RData")

##dim(total.log.norm.23.no.NA.R1D0.Reliability.no.undetected) # 7033 23



```